---
title: "How do fentanyl overdose rates vary with education levels across different states of the USA?"
author: "Julie Decraemer & Kenza Redwane"
format: 
  html: default
  pdf: default
output:
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage[margin=0.8in]{geometry}
  - \usepackage{longtable}
  - \usepackage{booktabs}
  - \usepackage{array}
filters:
  - _extensions/andrewheiss/wordcount/wordcount.lua
wordcount:
  display: true   
  output: document  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
#| message: false
here::i_am("DM-Project-4.Rproj")
library(here)
library(vroom)
library(ggplot2)
theme_set(theme_bw())
```

```{r}
#| message: false
drug_seizure <- vroom(here("drug-consumption.csv"))
```

```{r}
#| message: false
lethal_overdoses <- vroom(here("lethal-overdoses.csv"))
```

```{r}
#| message: false
education <- vroom(here("education.csv"))
```

```{r}
#| message: false
library(tidyverse)
library(readr)
library(knitr)
library(kableExtra)
education <- vroom(here("education.csv"))

education_clean <- education %>%
  slice(8:(n() - 3))%>%  

select(-c(3, 4, 6, 7, 9, 10, 12, 13, 15, 16, 18, 19, 21, 22, 24, 25, 
            27, 28, 30, 31, 33, 34, 36, 37, 39, 40, 42, 43)) %>%
rename(
  states = "Table 104.80. Percentage of persons 18 to 24 years old and age 25 and over. by educational attainment and state or jurisdiction: 2012 and 2022",
  less24years_highschool_2012 = "...2",
    less24years_highschool_2022 = "...5",
    years25_LessHighSchool_2012 = "...8",
    years25_LessHighSchool_2022 = "...11",
    years25_moreHighSchool_2012 = "...14",
    years25_moreHighSchool_2022 = "...17",
    years25_bachelor_higher_2012 = "...20",
    years25_bachelor_higher_2022 = "...23",
    years25_bachelor_2012 = "...26",
    years25_bachelor_2022 = "...29",
    years25_graduate_2012 = "...32",
    years25_graduate_2022 = "...35",
    years25_college_2012 = "...38",
    years25_college_2022 = "...41"
  ) 
```

```{=html}
<style>
caption {
  text-align: left;
}
</style>
```
```{r}
# Rename specific columns and remove column 12
education_clean <- education_clean %>%
  rename(
    # Specify new names for the columns
    `years25_atleast_highschool_2012` = 5,
    `years25_total_bachelor+_2012` = 6,
    `years25_bachelor_2012` = 7,
    `years25_graduate_2012` = 8,
    `years25_LessHighSchool_2022` = 9,
    `years25_atleast_highschool_2022` = 10,
    `years25_onlyhighschool_2022` = 11,
    `useless` = 12,
    `years25_total_bachelor+_2022` = 13,
    `years25_bachelor_2022` = 14,
    `years25_graduate_2022` = 15
  )

# Remove useless columns as well as columns on the year 2012
education_clean <- education_clean %>%
  select(-useless, -matches("2012"))

education_clean <- education_clean %>%
  # Remove everything after the comma (including the comma) and the periods
  mutate(across(-states, ~ gsub(",.*", "", .))) %>%  # Remove the comma and everything after it
  mutate(across(-states, ~ gsub("\\.", "", .))) %>%  # Remove the dots
  # Convert to numeric
  mutate(across(-states, ~ as.numeric(.)))

```

```{r}
state_name <- unique(education_clean$states) # Extraction of unique states' names

state_codes <- data.frame(
  state_name = state_name,
  state_code = c("USA", "AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC", "FL", 
                 "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", 
                 "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", "NM", 
                 "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", "SD", "TN", 
                 "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY", "OTHER JURISDICTIONS", "PR")
)

education_clean <- education_clean %>%
  left_join(state_codes, by = c("states" = "state_name")) %>%  # Jointure by states names
  mutate(states = state_code) %>%  # Replace names by the codes
  filter(states != "OTHER JURISDICTIONS") %>%  # Remove unwanted rows
  select(-state_code)
```

```{r}
#| message: false
drug_seizure <- drug_seizure %>%
  filter(FY == 2022)%>%
  rename(states = `Area of Responsibility`) 

```

```{r}
# Extract unique values from the 'states' column of the drug seizures data set
offices_ds <- unique(drug_seizure$states)

# Modifying the states values with the codes of the latest
ds_offices_codes <- data.frame(
  offices = offices_ds,
  codes = c(
    "GA", "MD", "IL", "TX", "CA", "FL", "LA", "NY", "INTL", "CA", "PR", "FL", 
    "MA", "NY", "MI", "WA", "OR", "TX", "TX", "CA", "AZ", "FL", "PR", "WA", "NY", 
    "MI", "ME", "WA", "VT", "TX", "TX", "CA", "TX", "TX", "TX", "CA", "AZ", "AZ", 
    "MT", "ND", "LA"
  )
)

# Column replacement
drug_seizure <- drug_seizure %>%
  left_join(ds_offices_codes, by = c("states" = "offices")) %>%  # Joining by 'offices'
  mutate(states = ifelse(!is.na(codes), codes, states)) %>%
  select(-codes)

```

```{r}
# Filter for only Fentanyl drug type
drug_seizure <- drug_seizure %>%
  filter(`Drug Type` == "Fentanyl")

```

```{r}
# Deleting useless variables (columns)
drug_seizure <- drug_seizure %>%
  select(-Region, -`Land Filter`, -Component)

# Exclude the state "INTL" as it isn't a state
drug_seizure <- drug_seizure %>%
  filter(states != "INTL")

# Agregating data by states
drug_seizure_clean <- drug_seizure %>%
  group_by(states) %>%
  summarise(
    total_events = sum(`Count of Event`, na.rm = TRUE),
    total_quantity = sum(`Sum Qty (lbs)`, na.rm = TRUE)
  ) %>%
  ungroup()

```

```{r}
# Add a USA row with the averages of the numeric columns
drug_seizure_clean <- drug_seizure_clean %>%
 filter(states != "USA") %>% # Remove the current USA row if it exists
  bind_rows(
    drug_seizure_clean %>%
      summarise(
        across(
          where(is.numeric), 
          ~ sum(.x, na.rm = TRUE) # Sum all numeric columns
        ),
        states = "USA" # Add the "USA" state name
      )
  )

# Adding the proportion variables
drug_seizure_clean <- drug_seizure_clean %>%
  mutate(
    total_events_proportion = round((total_events / total_events[states == "USA"]) * 100, 2),
    total_quantity_proportion = round((total_quantity / total_quantity[states == "USA"]) * 100, 2)
  )

```

```{r}
#| message: false
lethal_overdoses_clean <- lethal_overdoses %>%
  filter(YEAR == 2022) %>%  # Keep only data for the year 2022
  rename(states = STATE) %>%  # Rename the 'STATE' column to 'states'
  select(-URL, -YEAR) %>%  # Remove the 'URL' column
  mutate(states = if_else(states == "District of Columbia", "DC", states))

```

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
lethal_overdoses_clean <- lethal_overdoses_clean %>%
  bind_rows(
    lethal_overdoses_clean %>%
      summarise(
        across(
          .cols = where(is.numeric) & !c("DEATHS"),
          .fns = mean,
          na.rm = TRUE
        ),
        RATE = round(mean(RATE, na.rm = TRUE), 2),
        DEATHS = sum(DEATHS),
        states = "USA" # Add the observation `USA`
      )
  )
```

```{r}
# Perform the inner join on the three datasets using the 'states' variable
joined_dataset <- drug_seizure_clean %>%
  inner_join(education_clean, by = "states") %>%
  inner_join(lethal_overdoses_clean, by = "states")

```

## I. Research Question

The *fentanyl overdose crisis in the United States* represents a major \uline{public health emergency}, exacerbated by economic, social, and cultural factors. In *2022*, this crisis reached a \uline{critical point} and thus caught our attention for this research. That year, the United States recorded approximately 67,325 deaths due to fentanyl overdoses, accounting for nearly 70% of all overdose deaths that year, according to Watson.ch. At the same time, a significant increase in fentanyl production in China and its distribution by Mexican cartels worsened the crisis. This drug, often mixed with other substances to reduce costs, saw its consumption skyrocket. At the same time, post-COVID effects, such as depression and increased drug experimentation, worsened the crisis, particularly in poorer states, which were affected by high inflation and growing economic insecurity.

Focusing our analysis on 2022 ensures **consistent metrics across all states**. It allows for \uline{clearer} interpretation of results, avoiding complexities introduced by multi-year, time-based effects. In examining these patterns, **education** levels emerge as a potentially \uline{key factor in understanding regional disparities in this crisis}. Low education levels may limit access to information, increase risky behaviors, thus increasing the vulnerability of populations to fentanyl use and overdoses. According to Neuralword.com, about 35% of Americans aged 25 and older held a college degree in 2022, highlighting a potential educational gap influencing exposure to this crisis.

Thus, our **research question** is formulated as follows: *“How do fentanyl overdose rates vary with education levels across different states of the USA?”* More specifically, *“What is the relationship between education levels and fatal overdoses, and what regional disparities emerge from this?”*

Our objectives for this study would be:

-   To analyze the relationship between education levels and fentanyl overdose rates as well as with fentanyl consumption in different states.
-   To statistically examine regional disparities, whether regions with lower education levels are disproportionately affected by this crisis or not.

Thus, we have formulated the following hypotheses:

-   **States with lower education levels have higher rates of fentanyl overdoses and seizures.**
-   **Stronger negative relationship between fentanyl seizures and overdoses in less-educated states.**

## II. Description of Data Sets

### 1. Drug Seizures in the United States

```{r, warning = FALSE}
data_description_1 <- data.frame(
  Description = c(
    "Data on drug seizures between 2021 and 2024 in USA by Customs and Border Protection. It includes information on types, quantities (in pounds) and locations of drugs seized, used as an indicator of fentanyl consumption for illicit use. 
"),
  Variables = c(
    "16 observations and 5 columns (on 2022): `State`, character variable that presents postal code of the state of seizure. For each state, there is numeric variable, quantity of drugs seized, for each type of drugs. Finally, we have the number seizure events.
")
)
# Render the table
kable(data_description_1) %>%
  kable_styling(latex_options = c("scale_down", "striped"),
                full_width = FALSE,
                font_size = 12) %>%
  column_spec(1, width = "20em") %>%
  column_spec(2, width = "20em") %>%
  row_spec(0, bold = TRUE, color = "black")
```

### 2. Drug overdose mortality in the United States

```{r,warning = FALSE}
data_description_2 <- data.frame(
  Description = c(
    "Data on drug overdose mortality in the US from 1999 to 2022. It presents the total number of deaths for each state as well as the rate."),
  Variables = c(
    "52 observations and 3 variables during 2022.  The first column is States, character variable, that includes state postal code. Moreover, for each state, we have numeric variables that reflect number of Overdoses, and rate of deaths, per 100 000 habitants."
  )
)
# Render the table
kable(data_description_2) %>%
  kable_styling(latex_options = c("scale_down", "striped"),
                full_width = FALSE,
                font_size = 12) %>%
  column_spec(1, width = "20em") %>%
  column_spec(2, width = "20em") %>%
  row_spec(0, bold = TRUE, color = "black")
```

### 3. Education Levels in the United States

```{r,warning = FALSE}
data_description_3 <- data.frame(
  Description = c(
    "Data of educational attainment in US by state from 2012 to 2022. It contains total number of people attaining different levels of education for instance people completing high school, bachelor and graduate."),
  Variables = c(
    "53 observations and 8 variables during 2022.  The first column is States, character variable, that includes the state  postal code. For each state,  we have numeric variables expressing number of people under 24 completing highschool , total number of people of age 25+ who didn't completed secondary education, and those completing highschool, people with bachelor, those with at least a bachelor and finally those with graduate."
  )
)
# Render the table
kable(data_description_3, booktabs = TRUE, longtable = TRUE) %>%
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE,
    font_size = 12
  ) %>%
  column_spec(1, width = "20em") %>%
  column_spec(2, width = "20em") %>%
  row_spec(0, bold = TRUE, color = "black")  # Header styling
```

## III. A. Descriptive Analysis

### 1. Detailed Descriptive Analysis Table

```{r}
# Rename variables
joined_dataset <- joined_dataset %>%
  rename(
    total_seizure_events = total_events,
    total_quantity_seized = total_quantity,
    overdose_rate = RATE,
    overdose_deaths = DEATHS
  )%>%
  filter(states != "USA")
```

```{r, warning = FALSE}
#| message: false
#| cache: false

library(psych)
library(dplyr)
library(knitr)
library(kableExtra)

# Detailed Table of the Descriptive Analysis
numeric_dataset <- joined_dataset %>%
  select_if(is.numeric)  
summary_table <- describe(numeric_dataset) %>%
    mutate_if(is.numeric, ~ round(., 2))

filtered_summary_table <- summary_table[, c(3, 4, 5, 11, 12)]

# Render the numeric summary table
kable(filtered_summary_table, longtable = TRUE) %>%
  kable_styling(latex_options = c("striped"),  
                full_width = FALSE,
                font_size = 12) %>%
  column_spec(1, bold = TRUE, width = "12em") %>%
  column_spec(2, width = "8em") %>%
  column_spec(3, width = "5em") %>%
  column_spec(4, width = "5em") %>%
  column_spec(5,width = "3em") %>%
  column_spec(6,width = "3em") %>%
  row_spec(0, bold = TRUE, color = "black")
```

**Discussion**: <br>

-   For the variable **`total_quantity_seized`**, the *mean* is **1049.91**, indicating relatively high average quantity of drugs seized in the USA. The *standard deviation* of **2455.75** suggests great variability in seizures, showing that some seizures are much larger than the average. The *median*, at **16.01**, is significantly lower than mean, highlighting strong positive asymmetrical distribution, as evidenced by the *skew* of **1.96**. This implies several large seizures pulling average upwards. *Kurtosis*, at **2.27**, shows sharp distribution with heavy tails, suggesting that extreme values are frequent.

-   Concerning education, the variable **`years25_atleast_highschool_2022`** has *standard deviation* of **2825202.67**, indicating that majority of individuals have completed at least highschool with moderate variability. The *skew* of **0.15** indicates almost symmetrical distribution. For **`years25_bachelor_2022`**, the *standard deviation* of **3492597.44**, indicating fewer individuals have bachelor compared to highschool. The *kurtosis* of **-0.99** suggests slightly flatter distribution than normal. For **`years25_graduate_2022`**, the *standard deviation* of **3268291.44** reflects moderate variability. The *skew* of **0.79** highlights a concentration of lower values, indicating that higher diplomas are less common. Comparison between the averages for **`years25_bachelor_2022`** and **`years25_graduate_2022`** reflects the typical pyramid structure of education levels in a society.

-   Finally variable **`overdose_deaths`**, with *mean* of **3841.29** indicates an average number of fatal overdoses. The *standard deviation* of **2859.11** highlights high variability in overdose deaths, which is alarming. The relatively low *kurtosis*, about **0.29**, imply that the overdose problem is widespread rather than concentrated in a few extreme areas.

### 2. Median focused

To analyze the alignment between education, overdose, and seizure statuses, a composite metric is created for each category (education, overdose, and seizures) by calculating the row-wise mean of relevant variables. States are then classified as "success" or "failure" based on whether their composite score was above or below the national median. Here, the variables are classified as follow:

-   **Education Variables:** classified as "success" if their values are above national median and "failure" otherwise.

-   **Overdose Variables:** classified as "success" if their values are below national median and "failure" otherwise.

-   **Drug Seizure Variables:** classified as "success" if their values are below national median and "failure" otherwise.

```{r, warning = FALSE}
# Create composite scores and classify success/failure
success_status_by_median <- joined_dataset %>%
  mutate(
    # Composite education score and status
    composite_education = rowMeans(select(., years25_bachelor_2022, years25_graduate_2022), na.rm = TRUE),
    education_status = ifelse(composite_education > median(composite_education, na.rm = TRUE), "success", "failure"),
    
    # Composite overdose score and status
    composite_overdose = rowMeans(select(., matches("overdose")), na.rm = TRUE),
    overdose_status = ifelse(composite_overdose < median(composite_overdose, na.rm = TRUE), "success", "failure"),
    
    # Composite seizure score and status
    composite_seizure = rowMeans(select(., starts_with("total")), na.rm = TRUE),
    seizure_status = ifelse(composite_seizure < median(composite_seizure, na.rm = TRUE), "success", "failure")
  )
```

```{r, message =FALSE, echoes=FALSE, results='hide'}
# Cross-tabulation between education_status, overdose_status, and seizure_status
cross_tab <- table(
  Education = success_status_by_median$education_status,
  Overdose = success_status_by_median$overdose_status,
  Seizure = success_status_by_median$seizure_status
)

print(cross_tab)
```

```{r,warning = FALSE}
# Create a data frame for the table
contingency_data <- data.frame(
  Seizure_Status = c("Failure", "Failure", "Failure", "Failure", 
                     "Success", "Success", "Success", "Success"),
  Overdose_Status = c("Failure", "Failure", "Success", "Success",
                      "Failure", "Failure", "Success", "Success"),
  Education_Status = c("Failure", "Success", "Failure", "Success", 
                       "Failure", "Success", "Failure", "Success"),
  Count = c(2, 2, 2, 1, 
            2, 1, 1, 3)
)
# Render the contingency table
kable(contingency_data, caption = "Contingency Table for Success and Failure Across Metrics") %>%
  kable_styling(latex_options = c("scale_down", "striped"),
                full_width = FALSE,
                font_size = 12) %>%
  column_spec(1:3, bold = FALSE, width = "8em") %>%  
  column_spec(4, bold = TRUE, width = "6em") %>%  
  row_spec(0, bold = TRUE, color = "black")
```

\uline{Performance comparison accross variables:} States with better education metrics (success) are more likely to have lower overdose rates (success), especially when drug seizures fail. Education appears to be a strong protective factor against overdoses. Secondly, success in drug seizures doesn't consistently lead to lower overdose rates, highlighting that seizures alone are insufficient to address overdoses comprehensively.

### 3. Mean focused

<br> This descriptive analysis step categorizes states as "success" or "failure" based on their performance relative to the **mean** for three metrics: education, overdose rates, and drug seizures. Education is a "success" if above the mean, while overdose and drug seizures are "success" if below the mean. <br>

```{r, warning = FALSE}
success_status_by_mean <- joined_dataset %>%
  mutate(
    # Education variables: "success" if > mean, otherwise "failure"
    across(contains("2022"), ~ ifelse(. > mean(., na.rm = TRUE), "success", "failure")),

    # Overdose variables: "success" if < mean, otherwise "failure"
    across(matches("overdose_rate|overdose_deaths"), ~ ifelse(. < mean(., na.rm = TRUE), "success", "failure")),

    # Drug seizure variables: "success" if < mean, otherwise "failure"
    across(starts_with("total"), ~ ifelse(. < mean(., na.rm = TRUE), "success", "failure"))
  )

filtered_success_status <- success_status_by_mean[, c(1, 3, 12, 14)]

kable(head(filtered_success_status, 3), 
      caption = "Extract of Success and Failure Classification by Metric (Based on Mean)") %>%
  kable_styling(latex_options = c("scale_down", "striped"),
                full_width = FALSE,
                font_size = 12) %>%
  column_spec(1, bold = TRUE, width = "5em") %>% 
  column_spec(2:4, width = "11em") %>%  
  row_spec(0, bold = TRUE, color = "black")
```

```{r}
# Summarising success and failure counts
summary_success_status_by_mean <- success_status_by_mean %>%
  summarise(
    education_success = sum(across(contains("2022"), ~ . == "success"), na.rm = TRUE),
    education_failure = sum(across(contains("2022"), ~ . == "failure"), na.rm = TRUE),
    overdose_success = sum(across(matches("overdose_rate|overdose_deaths"), ~ . == "success"), na.rm = TRUE),
    overdose_failure = sum(across(matches("overdose_rate|overdose_deaths"), ~ . == "failure"), na.rm = TRUE),
    drug_seizure_success = sum(across(starts_with("total"), ~ . == "success"), na.rm = TRUE),
    drug_seizure_failure = sum(across(starts_with("total"), ~ . == "failure"), na.rm = TRUE)
  )
```

A proportional bar chart visualizes the distribution of "success" and "failure" across the three metrics (after summarization), using distinct colors to show relative performance.

```{r fig.align='center'}
# Prepare data for plotting directly from summary_success_status
plot_data <- data.frame(
  Category = c("Education", "Education", "Overdose", "Overdose", "Drug Seizure", "Drug Seizure"),
  Status = c("Success", "Failure", "Success", "Failure", "Success", "Failure"),
  Count = c(
    summary_success_status_by_mean$education_success,
    summary_success_status_by_mean$education_failure,
    summary_success_status_by_mean$overdose_success,
    summary_success_status_by_mean$overdose_failure,
    summary_success_status_by_mean$drug_seizure_success,
    summary_success_status_by_mean$drug_seizure_failure
  )
)

# Create the bar chart
ggplot(plot_data, aes(x = Category, y = Count, fill = Status)) +
  geom_bar(stat = "identity", position = "fill", width = 0.6) + # position = "fill" makes the bars proportional
  scale_fill_manual(values = c("Success" = "steelblue", "Failure" = "tomato")) + # Custom colors
  labs(
    title = "Proportion of Success and Failure by Category",
    x = "Category",
    y = "Proportion",
    fill = "Status"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent_format()) # Show proportions as percentages
```

-   **Comments:**

    The high proportion of "success" in the Drug Seizure category suggests that most states have relatively low drug seizure quantities and events compared to the mean. This could indicate effective enforcement or that fentanyl distribution isn't uniform across states. However, the correlation between drug seizures and overdoses needs further exploration—states with fewer seizures might still have high overdose rates due to the potency and availability of fentanyl.

    The Education category shows an almost equal distribution of "success" and "failure." This highlights significant variability in education levels across states, with some performing above the mean and others falling behind. For the research question, this variation in education levels could be crucial for understanding differences in overdose rates. States with higher education success may have populations better equipped to avoid risky behaviors or access preventive measures, potentially correlating with lower overdose rates.

    The Overdose category shows a higher proportion of "failure," indicating that many states have overdose rates or deaths exceeding the mean. This finding emphasizes that fentanyl overdoses remain a widespread problem, disproportionately affecting certain states. It suggests that efforts to mitigate overdoses need to focus on a majority of states where fentanyl use has escalated.

    The variability observed across education and overdose categories suggests a potential relationship between the two. States with high education failure (e.g., lower levels of high school or college attainment) might correspond to those with higher overdose failure, reflecting possible socioeconomic and educational disparities contributing to the fentanyl crisis. Conversely, states with education success might demonstrates better public health outcomes, including lower overdose rates. We will developp this later.

### 4. Relational Data Visualization

A- \uline{Overdoses and Seizures}

Examining the relationship between the quantity of fentanyl seized and the number of overdose deaths is critical because seizures are being used as a proxy (= an indicator) for drug consumption. By analyzing this relationship, we can explore whether higher quantities of fentanyl seized (and thus higher inferred consumption) are associated with increased overdose deaths. This analysis helps to validate the use of seizure data as an indicator of consumption trends.

**Relationship Between Total Quantity Seized and Quantity of Overdose:**

```{r, warning= FALSE, message =FALSE, echoes=FALSE, fig.align='center'}
library(ggplot2)
library(plotly)
# Fit the regression model
model <- lm(overdose_deaths ~ total_quantity_seized, data = joined_dataset)

# Predict values for the regression line
regression_line_1 <- data.frame(
  total_quantity_seized = seq(min(joined_dataset$total_quantity_seized, na.rm = TRUE),
                              max(joined_dataset$total_quantity_seized, na.rm = TRUE),
                              length.out = 100)
)
regression_line_1$overdose_deaths <- predict(model, newdata = regression_line_1)

# Create the ggplot with manual regression line
interactive_plot_1 <- ggplot() +
  geom_point(data = joined_dataset, aes(x = total_quantity_seized, 
                                        y = overdose_deaths, 
                                        text = paste("State: ", states,  "<br>Quantity seized: ", total_quantity_seized, 
                                                     "<br>Overdose Deaths: ", overdose_deaths)), 
             color = "blue") +
  geom_line(data = regression_line_1, aes(x = total_quantity_seized, y = overdose_deaths), 
            color = "red") +
  labs(
    x = "Total Quantity of Fentanyl Seized",
    y = "Number of Overdose Deaths (per 100,000)"
  )

interactive_plot_1
```

-   **Comments:**

    The graph shows a positive relationship between the total quantity of fentanyl seized and the number of overdose deaths per 100,000 people. This suggests that states with higher quantities of fentanyl seized tend to have higher overdose death rates. This implies that using drug seizure as a proxy of drug consumption is valid.

    To test the degree to which this conclusion is true, we perform a summary of this linear model:

```{r}
library(broom)
library(plotly)
```

```{r, message =FALSE, echoes=FALSE, warning = FALSE}
# Fit the linear model
model <- lm(overdose_deaths ~ total_quantity_seized, data = joined_dataset)

# Extract regression summary into a tidy format
regression_summary <- tidy(model)
model_stats <- glance(model)

# Create a summary table for the regression
summary_table <- data.frame(
  Metric = c("Intercept", "Slope (total_quantity_seized)", "R-squared", "Adjusted R-squared", "Residual Std. Error", "F-statistic", "p-value"),
  Value = c(
    regression_summary$estimate[1], 
    regression_summary$estimate[2], 
    model_stats$r.squared, 
    model_stats$adj.r.squared, 
    model_stats$sigma, 
    model_stats$statistic, 
    model_stats$p.value
  )
)

# Create a styled table for the regression summary
kable(summary_table, 
      caption = "Regression Summary: Relationship Between Total Quantity Seized and Overdose Deaths") %>%
  kable_styling(latex_options = c("scale_down", "striped"),
                full_width = FALSE,
                font_size = 12) %>%
  column_spec(1, width = "15em") %>%  
  column_spec(2, width = "10em") %>%
  row_spec(0, bold = TRUE, color = "black")  
```

The regression analysis reveals a **statistically significant positive relationship** between the quantity of fentanyl seized and overdose deaths. The coefficient for fentanyl seizures is 0.6547 (p = 0.03635), indicating that for every unit increase in fentanyl seized, the overdose death rate increases by 0.6547 per 100,000 people. The model explains 31.6% of the variance in overdose deaths (R-squared = 0.3162), highlighting a moderate relationship, though a significant amount of variation remains unexplained. While the model is statistically significant overall (F-statistic p = 0.03635), the residual standard error of 2461 indicates considerable variability in the data. **These findings suggest that higher fentanyl seizures may reflect greater fentanyl circulation, contributing to more overdose deaths.**

B- \uline{Education and Overdoses}

**Relationship Between Bachelor's Degree Number and Quantity of Overdose:**

```{r, warning= FALSE, fig.align='center'}
# Fit the regression model
model <- lm(overdose_deaths ~ years25_bachelor_2022, data = joined_dataset)

# Predict values for the regression line
regression_line_2 <- data.frame(
  years25_bachelor_2022 = seq(min(joined_dataset$years25_bachelor_2022, na.rm = TRUE),
                              max(joined_dataset$years25_bachelor_2022, na.rm = TRUE),
                              length.out = 100)
)
regression_line_2$overdose_deaths <- predict(model, newdata = regression_line_2)

# Create the ggplot with manual regression line
interactive_plot_2 <- ggplot() +
  geom_point(data = joined_dataset, aes(x = years25_bachelor_2022, 
                                        y = overdose_deaths, 
                                        text = paste("State: ", states, 
                                                     "<br>Bachelor's Degrees: ", years25_bachelor_2022, 
                                                     "<br>Overdose Deaths: ", overdose_deaths)), 
             color = "blue") +
  geom_line(data = regression_line_2, aes(x = years25_bachelor_2022, y = overdose_deaths), 
            color = "red") +
  labs(
    x = "Number of Bachelor's Degree",
    y = "Number of Overdose Deaths (per 100,000)"
  )

interactive_plot_2
```

-   **Comments:**

    The red trendline shows a clear negative slope, indicating an inverse relationship between the number of bachelor's and overdose deaths. States with higher numbers of bachelor's tend to have fewer overdose deaths, suggesting that higher education levels may correlate with better health outcomes or reduced drug-related mortality. The goal here is to check if this negative relationship is statistically significant, using the summary of the linear model.

```{r, message = FALSE, warning = FALSE}
# Fit the linear model
model <- lm(overdose_deaths ~ years25_bachelor_2022, data = joined_dataset)

# Extract regression summary into a tidy format
regression_summary <- tidy(model)
model_stats <- glance(model)

# Create a summary table for the regression
summary_table <- data.frame(
  Metric = c("Intercept", "Slope (years25_bachelor_2022)", "R-squared", "Adjusted R-squared", "Residual Std. Error", "F-statistic", "p-value"),
  Value = c(
    regression_summary$estimate[1], 
    regression_summary$estimate[2], 
    model_stats$r.squared, 
    model_stats$adj.r.squared, 
    model_stats$sigma, 
    model_stats$statistic, 
    model_stats$p.value
  )
)

kable(summary_table,
      caption = "Regression Summary: Relationship Between Bachelor's Degree Holders and Overdose Deaths") %>%
  kable_styling(latex_options = c("scale_down", "striped"),
                full_width = FALSE,
                font_size = 12) %>%
  column_spec(1, width = "15em") %>%  
  column_spec(2, width = "10em") %>%  
  row_spec(0, bold = TRUE, color = "black")  
```

The regression analysis indicates a **weak negative relationship** between **the number of bachelor’s degrees and overdose deaths**, as shown by the coefficient of -0.0001758, suggesting that for every increase of 10,000 bachelor’s degrees, the overdose death rate decreases by 1.758 per 100,000 people. However, this relationship isn't statistically significant at the 0.05 threshold, with a p-value of 0.098. The model explains only about 21% of the variance in overdose deaths (R-squared = 0.2104), highlighting its limited explanatory power. Additionally, the residual standard error of 2644 suggests substantial variability in the data. These results indicate that additional factors beyond education levels, such as economic or healthcare-related variables, likely play a significant role in influencing overdose deaths, and their inclusion may improve the model's effectiveness.

```{r}
joined_dataset_modified <- joined_dataset 

colnames(joined_dataset_modified)[colnames(joined_dataset_modified) == "years25_total_bachelor+_2022"] <- "years25_bachelorormore_2022"
```

C- \uline{Seizure and Education}

**Relationship Between Bachelor's Degree Number and the Quantity Seized:**

```{r, warning= FALSE, fig.align='center'}
# Fit the regression model
model <- lm(total_quantity_seized ~ years25_bachelor_2022, data = joined_dataset)

# Predict values for the regression line
regression_line_4 <- data.frame(
  years25_bachelor_2022 = seq(min(joined_dataset$years25_bachelor_2022, na.rm = TRUE),
                              max(joined_dataset$years25_bachelor_2022, na.rm = TRUE),
                              length.out = 100)
)
regression_line_4$total_quantity_seized <- predict(model, newdata = regression_line_4)

# Create the ggplot with manual regression line
interactive_plot_4 <- ggplot() +
  geom_point(data = joined_dataset, aes(x = years25_bachelor_2022, 
                                        y = total_quantity_seized, 
                                        text = paste("State: ", states, 
                                                     "<br>Bachelor's Degrees or more: ", years25_bachelor_2022, 
                                                     "<br>Overdose Deaths: ", total_quantity_seized)), 
             color = "blue") +
  geom_line(data = regression_line_4, aes(x = years25_bachelor_2022, y = total_quantity_seized), 
            color = "red") +
  labs(
    x = "Number of bechelor's Degree",
    y = "Total quantity of fentanyl seized"
  )

interactive_plot_4
```

```{r, warning=FALSE}
library(broom)
# Fit the linear model
model <- lm(total_quantity_seized ~ years25_bachelor_2022, data = joined_dataset)

# Extract regression summary into a tidy format
regression_summary <- tidy(model)
model_stats <- glance(model)

# Create a summary table for the regression
summary_table <- data.frame(
  Metric = c(
    "Intercept", 
    "Slope (years25_bachelor_2022)", 
    "R-squared", 
    "Adjusted R-squared", 
    "Residual Std. Error", 
    "F-statistic", 
    "p-value"
  ),
  Value = c(
    regression_summary$estimate[1],  # Intercept
    regression_summary$estimate[2],  # Slope
    model_stats$r.squared,           # R-squared
    model_stats$adj.r.squared,       # Adjusted R-squared
    model_stats$sigma,               # Residual Std. Error
    model_stats$statistic,           # F-statistic
    model_stats$p.value              # p-value
  )
)

# Create a formatted table
kable(summary_table, 
      caption = "Relationship Between Bachelor's Degree Holders and Total Quantity Seized") %>%
  kable_styling(latex_options = c("scale_down", "striped"), 
                full_width = FALSE, 
                font_size = 12) %>%
  column_spec(1, width = "15em") %>%  
  column_spec(2, width = "10em") %>%              
  row_spec(0, bold = TRUE, color = "black")    
```

-   **Comments:**

    The graph illustrates a **negative relationship** between **the number of bachelor's degrees** and the total quantity of fentanyl seized. This trend is further supported by the regression analysis results, where the coefficient for `years25_bachelor_2022` is -0.0001725, indicating that for every additional 10,000 bachelor's degrees, the total fentanyl quantity seized decreases by approximately 1.725 units. However, while the relationship is statistically significant at the 0.1 level (p-value = 0.0544), it isn't at the 0.05 threshold, limiting the confidence in this result. Additionally, the R-squared value of 0.2747 suggests that approximately 27.47% of the variance in fentanyl seizures is explained by the number of bachelor's degrees, highlighting some explanatory power but also leaving room for other contributing factors.

### 5. Regional Trends

This section focuses on cross-regional analysis as part of the descriptive study. Drug seizure data, available mostly for border states (e.g., Texas, California, Arizona), suggests trafficking patterns through these entry points. The analysis identifies regions with the highest overdose rates, education levels, and drug seizures, exploring factors like the southern region's proximity to Mexico, a major fentanyl source in 2022, and the northern region's higher education levels due to top universities. It also examines whether lower education levels correlate with greater opioid crisis impacts across regions.

A- \uline{Differences between regions}

We are first replicating `joined_dataset` but adding the variable `region` that will contain four observations: **north**, **south**, **west**, **midwest**.

```{r, warning=FALSE}
# Create a data frame to show states included in each region
region_states <- data.frame(
  Region = c("West", "North", "South", "Midwest"),
  States = c(
    "WA, CA, AZ, MT",
    "ME, NY, MA",
    "TX, LA, FL, GA, MD",
    "MI, IL"
  )
)

# Display the table using kable
library(kableExtra)

kable(region_states, 
      caption = "States Included in Each Region") %>%
  kable_styling(latex_options = c("scale_down", "striped"), 
                full_width = FALSE, 
                font_size = 12) %>%
  column_spec(1, width = "10em") %>%  
  column_spec(2, width = "20em") %>%             
  row_spec(0, bold = TRUE, color = "black")      
```

```{r, warning=FALSE}
joined_dataset_region_1 <- joined_dataset_modified %>%
  mutate(region = case_when(
    states %in% c("WA", "CA", "AZ", "MT") ~ "West",
    states %in% c("ME", "NY", "MA") ~ "North",
    states %in% c("TX", "LA", "FL", "GA", "MD") ~ "South",
    states %in% c("MI", "IL") ~ "Midwest"
  ))
joined_dataset_region <- joined_dataset_region_1 %>%
  select(states, region, everything())

joined_dataset_region <- joined_dataset_region %>%
  group_by(region) %>%
  summarise(
    total_seizure_events = sum(total_seizure_events, na.rm = TRUE),
    total_quantity_seized = sum(total_quantity_seized, na.rm = TRUE),
    less24years_highschool_2022 = sum(less24years_highschool_2022, na.rm = TRUE),
    years25_LessHighSchool_2022 = sum(years25_LessHighSchool_2022, na.rm = TRUE),
    years25_atleast_highschool_2022 = sum(years25_atleast_highschool_2022, na.rm = TRUE),
    years25_onlyhighschool_2022 = sum(years25_onlyhighschool_2022, na.rm = TRUE),
    years25_bachelorormore_2022 = sum(years25_bachelorormore_2022, na.rm = TRUE),
    years25_bachelor_2022 = sum(years25_bachelor_2022, na.rm = TRUE),
    years25_graduate_2022 = sum(years25_graduate_2022, na.rm = TRUE),
    overdose_rate = mean(overdose_rate, na.rm = TRUE),
    overdose_deaths = sum(overdose_deaths, na.rm = TRUE)
  )

filtered_joined_dataset_region <- joined_dataset_region[, c(1, 3, 9, 12)]

# Create a kable table for filtered_joined_dataset_region
kable(filtered_joined_dataset_region, 
      caption = "Summary of Regional Data") %>%
  kable_styling(latex_options = c("scale_down", "striped"), 
                full_width = FALSE, 
                font_size = 12) %>%
  column_spec(1, bold = TRUE, width = "8em") %>%  
  column_spec(2:ncol(filtered_joined_dataset_region), width = "10em") %>% 
  row_spec(0, bold = TRUE, color = "black")     
```

-   **Comments:**

    **Midwest region**: Despite having the second-lowest fentanyl seizures and the lowest overdose death rate per 100,000 inhabitants, it surprisingly has the lowest number of bachelor's degrees. This contradicts the hypothesis that lower overdose rates correlate with higher education levels. However, this region includes only two states, which might explain the lower number of bachelor’s degrees due to proportionality.
    
    **North region**: Comprising three states, it has the lowest fentanyl seizures and the second-lowest overdose death rate. Its education level is the second highest, aligning well with the hypothesis. Interestingly, the Midwest region seizes 95.43% more fentanyl but has 29.47% fewer overdose deaths, suggesting an inconsistent relationship between seizures and overdoses.
    
    **Southern region**: It shows the second-highest fentanyl seizures (915 lbs in 2022) and the highest overdose deaths (20,676). Education levels are also the highest among all regions, contradicting the hypothesis of a negative correlation between education and overdose rates. However, since this region has the most states, the higher data volume may inflate its metrics, reinforcing the need for correlation analysis.
    
    **West region**: With the highest fentanyl seizures, it also has the second-highest overdose deaths after the southern region. Its education level is similar to the north region, which aligns with earlier observations. Despite containing fewer states than the southern region, it still exhibits notable metrics for seizures and overdoses.

**We are now implementing an hypothesis testing accross regions using ANOVA summary, Tukey's test table, and boxplots.**

A- \uline{Overdose}

```{r, warning=FALSE, fig.align='center'}
# Perform ANOVA to test for differences in overdose rates across regions
anova_overdose <- aov(overdose_deaths ~ region, data = joined_dataset_region_1)

library(kableExtra)

# Create a summary table for ANOVA results
anova_overdose_summary <- summary(anova_overdose)[[1]]
anova_overdose_df <- data.frame(
  Source = rownames(anova_overdose_summary),
  Df = anova_overdose_summary[, "Df"],
  `Sum Sq` = anova_overdose_summary[, "Sum Sq"],
  `Mean Sq` = anova_overdose_summary[, "Mean Sq"],
  `F value` = anova_overdose_summary[, "F value"],
  `Pr(>F)` = anova_overdose_summary[, "Pr(>F)"]
)

# Render the ANOVA summary table for overdose rates
kable(anova_overdose_df, 
      caption = "ANOVA Results: Overdose Rates by Region") %>%
  kable_styling(latex_options = c("scale_down", "striped"), 
                full_width = FALSE, 
                font_size = 12) %>%
  column_spec(1, bold = TRUE, width = "5em") %>% 
  column_spec(2:ncol(anova_overdose_df), width = "6em") %>% 
  row_spec(0, bold = TRUE, color = "black")     

# Create a summary table for Tukey HSD test results
tukey_overdose <- TukeyHSD(anova_overdose)
tukey_overdose_df <- as.data.frame(tukey_overdose$region)


# Rename columns for better readability
colnames(tukey_overdose_df) <- c("Difference", "Lower Bound", "Upper Bound", "p-value")

# Display Tukey HSD results in a table
kable(tukey_overdose_df, 
      caption = "Tukey HSD Results: Overdose Rates by Region") %>%
  kable_styling(latex_options = c("scale_down", "striped"), 
                full_width = FALSE, 
                font_size = 12) %>%
  column_spec(1, bold = TRUE, width = "6em") %>%  
  column_spec(2:ncol(tukey_overdose_df), width = "5em") %>% 
  row_spec(0, bold = TRUE, color = "black")      

# Create a boxplot for visual representation
ggplot(joined_dataset_region_1, aes(x = region, y = overdose_deaths)) +
  geom_boxplot(fill = "skyblue", color = "darkblue") +
  labs(
    title = "Overdose Rates by Region",
    x = "Region",
    y = "Overdose Rate (per 100,000)"
  ) +
  theme_minimal()

```

-   **Comments:**

    The high p-values in the ANOVA results indicate that regional factors don't significantly explain the differences in overdose rates. Similarly, the Tukey HSD test showed no meaningful pairwise differences between regions, with confidence intervals including zero for all comparisons. The boxplot further supports these findings, as the distributions of overdose rates overlap considerably across regions, with no region showing distinctly higher or lower values. **These results suggest that overdose rates might not be primarily driven by regional characteristics.**

B- \uline{Seizures}

```{r, fig.align='center', warning=FALSE}
# Create a summary table for ANOVA results
anova_seizures <- aov(total_quantity_seized ~ region, data = joined_dataset_region_1)

anova_seizures_summary <- summary(anova_seizures)[[1]]
anova_seizures_df <- data.frame(
  Source = rownames(anova_seizures_summary),
  Df = anova_seizures_summary[, "Df"],
  `Sum Sq` = anova_seizures_summary[, "Sum Sq"],
  `Mean Sq` = anova_seizures_summary[, "Mean Sq"],
  `F value` = anova_seizures_summary[, "F value"],
  `Pr(>F)` = anova_seizures_summary[, "Pr(>F)"]
)

# Render the ANOVA results table for seizures
kable(anova_seizures_df, 
      caption = "ANOVA Results: Total Quantity Seized by Region") %>%
  kable_styling(latex_options = c("scale_down", "striped"), 
                full_width = FALSE, 
                font_size = 12) %>%
  column_spec(1, bold = TRUE, width = "5em") %>%  # Source column
  column_spec(2:ncol(anova_seizures_df), width = "6em") %>% # Other columns
  row_spec(0, bold = TRUE, color = "black")       # Header row

# Create a summary table for Tukey test results
tukey_seizures <- TukeyHSD(anova_seizures)
tukey_seizures_df <- as.data.frame(tukey_seizures$region)

# Rename columns for better readability
colnames(tukey_seizures_df) <- c("Difference", "Lower Bound", "Upper Bound", "p-value")

# Display Tukey test table
kable(tukey_seizures_df, 
      caption = "Tukey HSD Results: Total Quantity Seized by Region") %>%
  kable_styling(latex_options = c("scale_down", "striped"), 
                full_width = FALSE, 
                font_size = 12) %>%
  column_spec(1, bold = TRUE, width = "6em") %>%  
  column_spec(2:ncol(tukey_seizures_df), width = "5em") %>% 
  row_spec(0, bold = TRUE, color = "black")       

# Display the boxplot
ggplot(joined_dataset_region_1, aes(x = region, y = total_quantity_seized)) +
  geom_boxplot(fill = "lightgreen", color = "darkgreen") +
  labs(
    title = "Total Fentanyl Seized by Region",
    x = "Region",
    y = "Total Quantity Seized"
  ) +
  theme_minimal()

```

-   **Comments:**

    The analysis of fentanyl seizures across regions reveals interesting, though statistically non-significant, patterns. The ANOVA results (F(3,10) = 2.04, p = 0.172) suggest that there are no statistically significant differences in the total quantity of fentanyl seized across the four regions at the 5% significance level. However, the boxplot indicates that the \uline{West region exhibits the highest median and variability in the total fentanyl seizures compared to other regions}, suggesting that it might play a significant role as a primary entry or transit point for illicit drug trafficking. The Tukey HSD post-hoc test further supports the lack of statistically significant differences between region pairs.

C- \uline{Education}

```{r fig.align='center', warning=FALSE}
# Create a summary table for ANOVA results
anova_table <- summary(aov(years25_bachelor_2022 ~ region, data = joined_dataset_region_1))[[1]]
anova_df <- data.frame(
  Source = rownames(anova_table),
  Df = anova_table[, "Df"],
  `Sum Sq` = anova_table[, "Sum Sq"],
  `Mean Sq` = anova_table[, "Mean Sq"],
  `F value` = anova_table[, "F value"],
  `Pr(>F)` = anova_table[, "Pr(>F)"]
)

# Render the ANOVA table for years25_bachelor_2022
kable(anova_df, 
      caption = "ANOVA Results: Bachelor's Degree Holders by Region") %>%
  kable_styling(latex_options = c("scale_down", "striped"), 
                full_width = FALSE, 
                font_size = 12) %>%
  column_spec(1, bold = TRUE, width = "5em") %>%  
  column_spec(2:ncol(anova_df), width = "6em") %>% 
  row_spec(0, bold = TRUE, color = "black")

# Create a summary table for Tukey test results
tukey_test <- TukeyHSD(aov(years25_bachelor_2022 ~ region, data = joined_dataset_region_1))
tukey_df <- as.data.frame(tukey_test$region)

# Rename columns for better readability
colnames(tukey_df) <- c("Difference", "Lower Bound", "Upper Bound", "p-value")

# Display Tukey test table
kable(tukey_df, 
      caption = "Tukey HSD Test Results: Bachelor's Degree Holders by Region") %>%
  kable_styling(latex_options = c("scale_down", "striped"), 
                full_width = FALSE, 
                font_size = 12) %>%
  column_spec(1, bold = TRUE, width = "6em") %>%  
  column_spec(2:ncol(tukey_df), width = "5em") %>% 
  row_spec(0, bold = TRUE, color = "black")    

# Visualize the differences using a boxplot
ggplot(joined_dataset_region_1, aes(x = region, y = years25_bachelor_2022)) +
  geom_boxplot(fill = "#FFDAB9", color = "#D3D3D3" ) +
  labs(
    title = "Bachelor's Degree Holders by Region",
    x = "Region",
    y = "Percentage of Bachelor's Degree Holders"
  ) +
  theme_minimal()
```

-   **Comments:**

    The analysis of bachelor's degree holders across regions, as shown in the ANOVA table, indicates no statistically significant differences between regions (p = 0.414). This suggests that the proportion of bachelor's degree holders doesn't vary significantly across the Midwest, North, South, and West regions. The boxplot further supports this conclusion, as the distribution of bachelor's degree holders appears to overlap among regions, though the West region shows a wider range in bachelor's degree holders compared to others. The Tukey HSD test results confirm this lack of statistical significance, with all pairwise comparisons having high p-values. For instance, comparisons such as West-North (p = 0.734) and South-North (p = 0.967) indicate no meaningful differences in the proportion of bachelor's degree holders.

B- \uline{Correlation Analysis Accross States}

In this section of the regional trend analysis, we are using the Pearson Correlation to establish the relationship between our variables.

```{r, warning=FALSE, fig.align='center'}
# Select numeric variables for correlation
correl_dataset <- joined_dataset_region %>%
  select(overdose_deaths, total_quantity_seized, years25_bachelor_2022)

# Compute the correlation matrix
correlation_matrix <- cor(correl_dataset, use = "complete.obs", method = "pearson")

# Display the correlation matrix
correlation_table <- as.data.frame(as.table(correlation_matrix))

# Ensure Var1 and Var2 are characters, not factors
correlation_table <- correlation_table %>%
  mutate(
    Var1 = as.character(Var1),
    Var2 = as.character(Var2)
  )

# Filter out rows where Var1 equals Var2
correlation_table <- correlation_table %>%
  filter(Var1 != Var2) %>%
  # Remove duplicate pairs
  rowwise() %>%
  filter(Var1 < Var2) %>%
  ungroup()

kable(correlation_table, 
      caption = "Correlation Analysis Results") %>%
  kable_styling(latex_options = c("scale_down", "striped"), 
                full_width = FALSE, 
                font_size = 12) %>%
  column_spec(1, width = "10em") %>%  
  column_spec(2, width = "10em") %>%             
  column_spec(3, width = "8em") %>%             
  row_spec(0, bold = TRUE, color = "black")      
```

```{r, message=FALSE}
library(reshape2)

# Reshape correlation matrix for plotting
correlation_melt <- melt(correlation_matrix)

# Plot heatmap
ggplot(correlation_melt, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  labs(
    title = "Correlation Heatmap",
    x = "Variables",
    y = "Variables",
    fill = "Correlation"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

-   **Comments:**

    The two representations displaying correlations between our variables show three quite absurd contradictions. The first one is a **strong positive correlation between `overdose_deaths` and `years25_bachelor_2022` (0.88)**. This indicates that regions with a higher proportion of individuals holding at least a bachelor also tend to have higher overdose death rates. While surprising, this could suggest that higher education isn't a protective factor against overdoses or that other confounding factors, such as urbanization, play a role. The second one is a **moderate positive correlation between `overdose_deaths` and `total_quantity_seized` (0.37)**. This suggests that regions with higher drug seizure quantities also experience more overdose deaths. This relationship is expected and was already showed in the previous part of the descriptive analysis, as higher quantities of seized drugs could indicate higher levels of drug trafficking or consumption in these regions. Finally, a **weak positive correlation** between total_quantity_seized and years25_bachelor_2022 (0.09). This time it suggests little to no relationship between the quantity of drugs seized and the education level. This may imply that education levels don't directly influence the volume of drug trafficking. 

    In the previous part of our analysis, we found a negative relationship between the number of bachelor's degrees and overdose deaths in the regression model. The fact that this is contradictory with the present results of the correlation could be due to differences in the number of states within each region. For example, regions like the South and West, which include more states, may have a stronger influence on the overall trend compared to smaller regions like the North and Midwest.

    By comparing both approaches, we can conclude that the negative relationship in the regression model might be partially driven by regional aggregation and averaging effects, as well as differences in the number of contributing states per region, as said before. The correlation analysis provides complementary perspective.

```{r, message=FALSE, fig.align='center'}
# Pairwise scatter plots
library(GGally)

ggpairs(correl_dataset, title = "Pairwise Correlations Between Variables")
```

-   **Comments:**

    This pairwise correlation plot provides a deeper visual understanding of the relationships between the variables. We notice that while the correlation between the number of bachelor and overdose deaths is strong and positive (0.88), the scatterplot reveals **non-linear patterns, suggesting the relationship might not be perfectly linear across all regions**. Similarly, the relationship between total fentanyl seized and overdose deaths shows a moderate positive correlation (0.37), but the scatterplot indicates **variability and potential outliers influencing the trend**. Interestingly, the relationship between total fentanyl seized and the number of bachelors (0.086) is very weak, and the scatterplot highlights a lack of clear pattern, which aligns with the correlation coefficient. **This shows that fentanyl seizures aren't strongly associated with education levels when aggregating the states into regions, possibly due to factors like trafficking routes, which might not directly depend on the regional educational background.**

## III. B. Predictive Analysis

### 1. Regression model and estimations

Since our reasearch question is causal, we'll need to characterize and predict  links between our variables. We'll investigate those links through estimation of our regression model: 

-   [**Dependent variable (Y) :**]{.underline}

**- Overdose_Deaths:** metric  quantifies number of fatal overdoses

-   [**Independent variables (X):**]{.underline}

**- X1 = Years25_LessHighSchool_2022 :** expected to have positive relationship with overdose, as lower education levels may correlate with higher risks of substance misuse.

Same for **- X2 = years25_LessHighSchool_2022:** and **X3 = years25_onlyhighschool_2022**​

**- X4 = ​years25_bachelor_2022​** supposed to have negative relationship with overdose deaths, since higher education levels may correlate with better health outcomes.

**- X5 = years25_graduate_2022**  we assume negative relationship with overdose deaths, reflecting potential protective effects of higher education.

**- X6 = total_seizure_events​** We expect positive relationship since higher number of seizure events imply higher consumers of drugs which potentially leads to overdoses

**- X7 = Total_Quantity_Seized** supposed to have positive relationship with overdose, since high quantity seized means higher consumption of drugs, leading to higher deaths

Hence, we get *multiple linear regression model* such that:

```{r, echo=FALSE, results='hide'}
# Linear regression model
library(broom)
regression_model <- lm(overdose_deaths ~ less24years_highschool_2022 + 
                             years25_LessHighSchool_2022 + 
                             years25_onlyhighschool_2022 +
                             years25_bachelor_2022 + 
                             years25_graduate_2022 + 
                             total_seizure_events + 
                             total_quantity_seized,  
                data = joined_dataset)

tidy(regression_model) 
```

**Overdose_Deaths i** =$\beta0$ + $\beta1$\*​(less24years_highschool_2022​) + $\beta2$​\*(years25_LessHighSchool_2022​) +$\beta3$​ \* ​(years25_onlyhighschool_2022​) + $\beta4$​\*​(years25_bachelor_2022​)+$\beta5$​\*​(years25_graduate_2022​)+ $\beta6$​\*​(total_seizure_events​) + $\beta7$​\*​(total_quantity_seized​)+$\epsilon𝑖$​

Where:

-   $\beta1$ : Intercept.
-   $\beta1$,$\beta2$... $\beta7$: Coefficients associated with explanatory variables.
-   $\epsilon 𝑖$: Random error term, assumed to follow a normal distribution 𝑁(0,𝜎2)

**Test Hypothesis**

-   [Null hypothesis (H0) :]{.underline}

    -   **H0 :** $\beta0$ = 0

    -   There is no statistically significant relationship between the independent variable 𝑋𝑖 and the number of overdose deaths. This means that 𝑋𝑖 doesn't explain variations in overdose deaths.

-   [Alternative hypothesis (H1):]{.underline}

    -   **H1 :**    $\beta𝑖$ ≠ 0

    -   There is a statistically significant relationship between the independent variable 𝑋𝑖 and the number of overdose deaths. In other words, X𝑖 helps explain variations in overdose deaths

**Our Assumptions:**

The multiple linear regression model take into consideration some assumptions:

1.  Relationship between the independent variables (education levels) and dependent variable (overdose deaths) is [*linear*]{.underline}.

2.  Observations (states) are [*independent*]{.underline},  corresponds to index i.

3.  Variance of the residuals is [*constant*]{.underline} across all of the independent variables.

4.  Residuals of the model are [*normally distributed*]{.underline}.

5.  The independent variables (specially education levels) aren't [*highly*]{.underline}

    [*correlated*]{.underline} with each other.

**Estimation of the multiple linear regression linear:**

We use ordinary least squares (OLS) method to estimate the model coefficients.

```{r, warning=FALSE}
library(knitr)
library(kableExtra)
# Extract the model
summary_table <- summary(regression_model)$coefficients
colnames(summary_table) <- c("Estimate", "Std. Error", "t-value", "Pr")

# We convert the table into HTML format using kableExtra
kable(summary_table, 
      caption = "Results of the linear regression") %>%
  kable_styling(latex_options = c("scale_down", "striped"), 
                full_width = FALSE, 
                font_size = 12) %>%
  column_spec(1, bold = TRUE, width = "12em") %>%  # Coefficient names column
  column_spec(2, width = "8em") %>%               # Estimate column
  column_spec(3, width = "6em") %>%               # Std. Error column
  column_spec(4, width = "5em") %>%               # t-value column
  column_spec(5, width = "5em") %>%               # Pr column
  row_spec(0, bold = TRUE, color = "black")       # Header row
```

-   **i) Statistical Significance Analysis of Coefficients**

    \- The **intercept** coefficient estimates $\beta0$ = 7982.457 reflects positive value, suggesting that when all independent variables are zero,number of overdose deaths is 7982. Since p=0.558, so the intercept isn't significant, since it  is unlikely that education would be zero in real world

    \- The variable **less24years_highschool_2022** has coefficient of $\beta1$ = 0.000177, and p-value **0.558507**. This explains that an additional person having completed highschool implies an increase of 0.000177 in number of overdose .The effect of this variable on overdose deaths appears to be very small and statistically insignificant. Therefore, for this model, the proportion of young people under 24 without a highschool diploma isn't a major explanatory factor.

    \- The coefficient of the variable **years25_LessHighSchool_2022** is $\beta2$ = 0.0007049 with a p-value of **0.08** (close to significance at the 0.05 level). Indeed, the variable has moderate positive effect on overdose deaths. Although not strictly significant, it indicates that high proportions of people aged 25 and over without a highschool diploma are associated with an increase in overdose deaths.

    \- The variable **years25_onlyhighschool_2022** has an estimate of $\beta3$= -0.0003639 and **p-value** of **0.33** (not significant). Hence, this variable shows a non-significant negative effect. This suggests that a higher proportion of young people with only a highschool diploma (and nothing beyond) is associated with slight reduction in overdose deaths, but this result isn't conclusive.

    \- The estimate of **years25_bachelor_2022** is $\beta4$= -0.0000808 with a **p-value : 0.49** (not significant). The proportions of people with a university degree (bachelor) also appear to have a non-significant negative effect on overdose deaths.This may be explained by the fact that higher education offers greater protection against overdose deaths, but the result remains weak.

    \- The variable **years25_graduate_2022** has an estimate of $\beta5$ = -0.0000497 and a p-value of **0.88** (not significant). The proportions of graduates have no significant effect on overdose deaths in this model.

    \- The estimate of the variable **total_seizure_events** is $\beta6$ = -7.6052413 with p-value of **0.71** (not significant). Thus, the total number of seizure events appears to have a (small) negative effect on overdose deaths, but this result isn't statistically significant.

    \- The variable **total_quantity_seized** has an estimate $\beta7$ = 0.3498855 and a p-value of **0.80** (not significant). Thus, the total quantity of substances seized has a non-significant positive effect on overdose deaths.This reflects the fact that regions where more drugs are seized are also those where access to substances is higher with high overdose deaths, but this relationship remains non-significant.

-   **ii) Interpretation of the main results:**

    None of the model variables is statistically significant at the 0.05 level, although **years25_LessHigh School_2022** approaches marginal significance.This could be due to small sample size or collinearity between explanatory variables.\

    Moreover, variables associated with higher levels of education (bachelor, graduate) have negative coefficients, which is consistent with the idea that education may play a protective role. It suggests that individuals with greater educational attainment tend to be more aware of the risk of fentanyl and drugs consumption. On the other hand, higher educated people have better access to health care, and improved health literacy\

    Variables associated with lower levels of education (less than high school) tend to have positive coefficients, implying a potential association with higher overdose rates.

### 2. Model fit analysis

**1. Verification of our assumptions**

We start by checking that the model respects the assumptions of multiple linear regression.

-   [**a. Linearity**]{.underline}:

In order to check the linearity, we will plot residuals against predicted values:

```{r fig.align='center'}
# we calculate the residual values
residuals <- resid(regression_model)

# Predicted values
fitted_values <- fitted(regression_model)

# graphics with residuals
plot(fitted_values, residuals,
     xlab = "Values predicted",
     ylab = "Residuals",
     main = "Graphics of the residuals with errors",
     pch = 19, col = "blue")

# we add an horizontal line y = 0
abline(h = 0, col = "red", lwd = 2)

# we add a curve in order to represent an estimation of errors. 
lines(lowess(fitted_values, residuals), col = "green", lwd = 2)

# Legend
legend("topright", legend=c("Residuals", "Error estimated"), col=c("blue", "green"), pch=c(19, NA), lty=c(NA, 1))
```

-   **Comments:**

    The green line shows a slightly non-linear trend (curved shape). This may indicate that the relationship between our independent variables and the overdoses deaths isn't [strictly linear]{.underline}.

    Moreover, residuals appear to vary unequally around the red line. For instance, residuals on the left (low predicted values) and right (high predicted values) appear to be more dispersed. This could indicate [heteroscedasticity]{.underline}.

Now we correct the violation of the first assumption, one alternative is to use the spline since quadratic form and log didn't solve the linearity problem in our model.

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE, fig.align='center'}
library(splines)

# Use the splines
spline_model <- lm(overdose_deaths ~ bs(years25_LessHighSchool_2022, df = 3) +
                   bs(years25_onlyhighschool_2022, df = 3) + 
                   bs(years25_bachelor_2022, df = 3) + 
                   bs(years25_graduate_2022, df = 3) + 
                   bs(total_seizure_events, df = 3) + 
                   bs(total_quantity_seized, df = 3), 
                   data = joined_dataset)
summary(spline_model)

# Verify the residuals
residuals <- resid(spline_model)
fitted_values <- fitted(spline_model)

plot(fitted_values, residuals,
     xlab = "Values predicted",
     ylab = "Residuals",
     main = "Graphic of residuals(Splines)",
     pch = 19, col = "blue")
abline(h = 0, col = "red", lwd = 2)
lines(lowess(fitted_values, residuals), col = "green", lwd = 2)
```

-   [**b. Independence**]{.underline}

We make sure that the residuals aren't correlated in our model, thus we use Durbin-Watson test:

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(lmtest)

# Test of Durbin-Watson
dw_test <- dwtest(regression_model)
# The final result
cat("Durbin-Watson Test Result is DW =", round(dw_test$statistic, 4),"and p-value =", round(dw_test$p.value, 4),"\n" )
cat("Thus, Residuals aren't significantly autocorrelated.\n")
```

-   **Comments:**

    So, we don't reject the null hypothesis of no autocorrelation of the residuals, as p-value is well above the usual level of 0.05. There is no significant evidence of positive autocorrelation in the residuals of the regression model, suggesting that the observations are independent of each other.

-   [**c. Homoscedasticity**]{.underline}

Let's verify that the variance of residuals is constant using Breusch-Pagan test

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Test of Breusch Pagan
bp_test_result <- bptest(regression_model)
# The final result
cat("The studentized Breusch-Pagan test result is BP =", round(bp_test_result$statistic, 4),"and p-value =", round(bp_test_result$p.value, 4),"\n" )

```

-   **Comments:**

    We don't reject the null hypothesis of homoscedasticity, since p-value is above the usual 0.05 level. There is no significant evidence of heteroscedasticity in the residuals of the regression model, implying that the variance of the residuals is constant across the predicted values and that the model homoscedasticity assumption is respected.

-   [**d. Normality of distribution**]{.underline}

We check normality of distribution using Shapiro-Wilk test. Test's null hypothesis means that data are normally distributed (the alternative hypothesis is inverse).

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# We calculate the residuals
residuals <- resid(regression_model)

# the Shapiro-Wilk test
shapiro_test_result <- shapiro.test(residuals)
cat("Shapiro-Wilk normality test is W =", round(shapiro_test_result$statistic, 4),"and p-value =", round(shapiro_test_result$p.value, 4),"\n" )

```

-   **Comments:**

    We don't reject the null hypothesis that the residuals follow a normal distribution, as the p-value is above the usual 0.05 level. There is insufficient evidence to conclude that the residuals of the regression model aren't normally distributed, suggesting that this hypothesis is met in our analysis.

In order to visualize it, we draw QQ plot

```{r, fig.align='center'}
#Graphic Q-Q
qqnorm(residuals)
qqline(residuals, col = "red")  # we add a red line as a reference  
```

The graph shows Q-Q Plot to check the normality of the data. The points generally follow red diagonal line, indicating close-to-normal distribution, although there are deviations at the ends (the tails), suggesting possible extreme values or a slight deviation from normality.

The Density of the normalized residuals would be:

```{r, fig.align='center'}
# we extract residuals in our model  
residuals <- resid(regression_model)

# we normalize residuals into stardardized residuals
std_residuals <- scale(residuals)

# Density of standardized residuals
plot(density(std_residuals), 
     main = "Density of Normalized Residuals", 
     xlab = "Standardized Residuals (X)", 
     ylab = "Density", 
     col = "black", 
     lwd = 2)

# adding vertical lines forconfidence interval  (95%)
abline(v = c(-1.96, 1.96), col = "black", lty = "dashed")

# adding shadow areas in the tails distribution
x <- seq(-4, 4, by = 0.001)
y <- dnorm(x)

# Area left of -1.96
polygon(c(x[x <= -1.96], -1.96), c(y[x <= -1.96], 0), col = "gray", border = NA)

#Area right of -1.96
polygon(c(x[x >= 1.96], 1.96), c(y[x >= 1.96], 0), col = "gray", border = NA)

# adding legends
legend("topright", 
       legend = c("Density of residuals", "Interval of confidence (95%)"), 
       col = c("black", "black"), 
       lty = c(1, 2), 
       lwd = c(2, 1)) 
```

Using the Density of Normalized Residuals test, the hypotheses can be evaluated. The results show that the coefficients of the explanatory variables aren't significant (p \> 0.05) for most variables, indicating that the null hypothesis (H0) cannot be rejected. This implies that the independent variables don't show a statistically significant relationship with overdose deaths.

**2. Statistic Tests**

Now, we proceed with statistical tests to assess relationships between variables.

-   [**Chi-squared test χ2**]{.underline}

Since we have numerical variables, we should categorize them in order to use the Chi test, using quartiles.

```{r}
# Categorization of years25_LessHighSchool_2022
joined_dataset$less_highschool_category <- cut(joined_dataset$years25_LessHighSchool_2022, 
                                               breaks = quantile(joined_dataset$years25_LessHighSchool_2022, probs = seq(0, 1, 0.25), na.rm = TRUE), 
                                               labels = c("Low", "Medium", "High", "Very High"), 
                                               include.lowest = TRUE)

# Categorization of other Education Variables
joined_dataset$highschool_category <- cut(joined_dataset$years25_onlyhighschool_2022, 
                                          breaks = quantile(joined_dataset$years25_onlyhighschool_2022, probs = seq(0, 1, 0.25), na.rm = TRUE), 
                                          labels = c("Low", "Medium", "High", "Very High"), 
                                          include.lowest = TRUE)

joined_dataset$bachelor_category <- cut(joined_dataset$years25_bachelor_2022, 
                                        breaks = quantile(joined_dataset$years25_bachelor_2022, probs = seq(0, 1, 0.25), na.rm = TRUE), 
                                        labels = c("Low", "Medium", "High", "Very High"), 
                                        include.lowest = TRUE)

joined_dataset$graduate_category <- cut(joined_dataset$years25_graduate_2022, 
                                        breaks = quantile(joined_dataset$years25_graduate_2022, probs = seq(0, 1, 0.25), na.rm = TRUE), 
                                        labels = c("Low", "Medium", "High", "Very High"), 
                                        include.lowest = TRUE)

# Categorization of overdoses
joined_dataset$overdose_category <- cut(joined_dataset$overdose_deaths, 
                                        breaks = c(200, 1000, 3000, 5000, 8000, 11000), 
                                        labels = c("200-1000", "1001-3000", "3001-5000", "5001-8000", "8001-11000"), 
                                        include.lowest = TRUE)
```

Once all variables categorized, we create contingency table to explore the relationship between each categorical variable and overdoses_deaths variables.

```{r, warning = FALSE}
library(kableExtra)

# We create unique contingency table for all the variables of educations VS Overdose deaths  
contingency_table <- data.frame(
  "Overdose Category" = joined_dataset$overdose_category,
  "Less Highschool" = joined_dataset$less_highschool_category,
  "Highschool" = joined_dataset$highschool_category,
  "Bachelor" = joined_dataset$bachelor_category,
  "Graduate" = joined_dataset$graduate_category
)

# Initialisation of the results
results <- data.frame(
  Variable = c("Less Highschool", "Highschool", "Bachelor", "Graduate"),
  Chi_Square = numeric(4),
  DF = integer(4),
  P_Value = numeric(4)
)

# For each variable of education we do the Chi squared test 
for (i in 2:ncol(contingency_table)) {
  contingency <- table(contingency_table[, i], contingency_table$Overdose.Category)
  chisq_test <- chisq.test(contingency)
  
  results$Chi_Square[i - 1] <- chisq_test$statistic
  results$DF[i - 1] <- chisq_test$parameter
  results$P_Value[i - 1] <- chisq_test$p.value
}

# Results using Kable
results %>%
  kable(
    caption = "Chi-Squared Test Results for Education Variables vs Overdose Categories",
    col.names = c("Variable", "Chi-Square", "Degrees of Freedom (DF)", "p-Value")
  ) %>%
  kable_styling(
    latex_options = c("scale_down", "striped"),
    full_width = FALSE,
    font_size = 12
  ) %>%
  column_spec(1, bold = TRUE, width = "10em") %>%
  column_spec(2, width = "12em") %>%
  column_spec(3, width = "8em") %>%
  column_spec(4, width = "8em") %>%
  row_spec(0, bold = TRUE, color = "black")
```

-   **Comments:**

    The results of Chi-square tests indicate absence of statistically significant relationships between levels of education (Less Highschool, Highschool, Bachelor, Graduate) and categories of overdose deaths in the different states. The p-values for all educational variables (0.229, 0.158, 0.205, and 0.517 respectively) are above the level of 0.05, suggesting that we cannot reject the null hypothesis H0 of independence between these variables. **In other words, differences in the proportions of overdose deaths don't vary significantly according to education levels.**

    Regarding the State variable, the Chi-square test shows a p-value of 0.327, also above the 0.05 significance level. Again, there is no significant relationship between states and overdose categories. This observation could indicate that regional disparities aren't sufficiently marked, or that differences in overdoses are influenced by other factors not included in this analysis.

    These results raise questions about the real influence of education levels on the fentanyl overdose crisis. Although education is perceived as protective factor, these results don't show a statistically significant association. This could be explained by some factors such as limited data but also confounding factors, for instance, health access, economic conditions...

<!-- -->

-   [**The Student test**]{.underline}

Our second test, Student test, allows us to measure the difference between the observed mean and $\mu$ adjusted by standard deviation and sample size. Here, we do the 2 Sample t-test, which essentially compares overdose deaths and highschool category. Our hypothesis is as follows:

**Null hypothesis**(H0): The means of overdose deaths for the “Low” and “Very High” groups (based on highschool_category) are equal. In other words, H0 $\mu$ Low = $\mu$ Very High

**Alternative hypothesis**(H1): The means of overdose deaths for the two groups are different. Therefore: H1: $\mu$ Low $≠$ $\mu$ Very High

```{r, warning=FALSE}
library(kableExtra)
t_test_results <- data.frame(
  Statistic = c("t-statistic", "Degrees of Freedom (df)", "P-Value", "Confidence Interval (95%)", "Mean (Low)", "Mean (Very High)"),
  Value = c(1.4454, 3.6066, 0.2293, "[-3171.294, 9473.294]", 4723, 1572)
)

t_test_results %>%
  kable(
    caption = "Welch Two Sample t-Test Results: Highschool Education Category vs Overdose Deaths",
    col.names = c("Statistic", "Value"), 
    align = "l", 
    escape = FALSE 
  ) %>%
  kable_styling(
    latex_options = c("scale_down", "striped"), 
    full_width = FALSE,
    font_size = 12
  ) %>%
  column_spec(1, bold = TRUE, width = "12em") %>%
  column_spec(2, width = "15em") %>%
  row_spec(0, bold = TRUE, color = "black")
```

-   **Comments:**

    Welch's t-test compares overdose deaths between the "Low" and "Very High" groups according to secondary education level. The results show a higher average for the "Low" group (4723 deaths) compared to the "Very High" group (1572 deaths). However, this difference isn't statistically significant (p=0.2293), as also indicated by the wide confidence interval (-3171.294, 9473.294), which includes 0. It reflects high variability in data and limited sample, reducing the ability to detect a significant difference.

Graphically we get:

```{r, warning = FALSE, fig.align='center'}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Filter the data for the relevant groups
low_group <- joined_dataset %>% filter(highschool_category == "Low") %>% select(overdose_deaths)
very_high_group <- joined_dataset %>% filter(highschool_category == "Very High") %>% select(overdose_deaths)

# Combine the groups into a single dataframe with an indicator
plot_data <- data.frame(
  overdose_deaths = c(low_group$overdose_deaths, very_high_group$overdose_deaths),
  group = c(rep("Low", nrow(low_group)), rep("Very High", nrow(very_high_group)))
)

# Calculate means for each group
mean_low <- mean(low_group$overdose_deaths, na.rm = TRUE)
mean_very_high <- mean(very_high_group$overdose_deaths, na.rm = TRUE)

# Create the density plot
ggplot(plot_data, aes(x = overdose_deaths, fill = group)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = mean_low, color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = mean_very_high, color = "blue", linetype = "dashed", size = 1) +
  annotate("text", x = mean_low, y = 0.0001, label = paste("Mean (Low) =", round(mean_low, 2)),
           color = "red", hjust = -0.1) +
  annotate("text", x = mean_very_high, y = 0.0001, label = paste("Mean (Very High) =", round(mean_very_high, 2)),
           color = "blue", hjust = 1.2) +
  labs(
    title = "Density Plot of Overdose Deaths by Highschool Education Level",
    x = "Overdose Deaths",
    y = "Density"
  ) +
  scale_fill_manual(values = c("red", "blue")) +
  theme_minimal() +
  theme(legend.title = element_blank(), legend.position = "top")

```

-   **Comments:**

    Density graph clearly illustrates differences in the distribution of overdose deaths between the "Low" and "Very High" groups. The distribution of deaths for the "Low" group is wider and shows a higher mean (4723) than that of the "Very High" group (1572). However, superposition of distributions indicates high variability in both groups

**3. Adjustment measures**

```{r, warning=FALSE}
# Extract the model summary
model_summary <- summary(regression_model)

# Extract key statistics
residual_se <- model_summary$sigma  # Residual standard error
r_squared <- model_summary$r.squared  # Multiple R-squared
adjusted_r_squared <- model_summary$adj.r.squared  # Adjusted R-squared
f_statistic <- model_summary$fstatistic  # F-statistic
f_value <- f_statistic[1]  # F-statistic value
df1 <- f_statistic[2]  # Degrees of freedom for numerator
df2 <- f_statistic[3]  # Degrees of freedom for denominator
p_value <- pf(f_value, df1, df2, lower.tail = FALSE)  # p-value for F-statistic

# Create a summary table
summary_table <- data.frame(
  Metric = c("Residual Standard Error", 
             "Multiple R-squared", 
             "Adjusted R-squared", 
             "F-statistic",
             "p-value"),
  Value = c(round(residual_se, 4), 
            round(r_squared, 4), 
            round(adjusted_r_squared, 4), 
            paste(round(f_value, 2), "on", df1, "and", df2, "DF"),
            format(p_value, scientific = TRUE))
)

# Render table
library(kableExtra)
summary_table %>%
  kable(
    caption = "Model Fit Statistics",
    col.names = c("Metric", "Value"), 
    align = "l", 
    escape = FALSE 
  ) %>%
  kable_styling(
    latex_options = c("scale_down", "striped"),
    full_width = FALSE,
    font_size = 12
  ) %>%
  column_spec(1, bold = TRUE, width = "15em") %>%
  column_spec(2, width = "20em") %>%
  row_spec(0, bold = TRUE, color = "black")

```

-   **Comments:**

    The statistics of our regression model show a high residual standard error (2187.4975), indicating significant variability in the model's predictions. The multiple R-squared of 0.7298 suggests that 72.98% of the variance in overdose deaths is explained by the included independent variables. However, R-squared adjustment, at 0.4146, reveals that after adjusting for the number of predictors, explanatory power of the model is reduced, thus highlighting a potential over-fitting or low robustness. F-statistic analysis (2.32 with 7 and 6 degrees of freedom) isn't significant (p = 0.1629627), indicating that the model as a whole doesn't significantly explain the variance in deaths. In other words, although some relationships may exist, the model isn't entirely reliable and may require improvements, such as the inclusion of additional variables.

## IV. Conclusion

Our research aimed to explore the relationship between education levels and fentanyl overdose rates across U.S. states while examining regional disparities. We provide answers to our initial objectives as follows:

The findings indicate that \uline{states with lower education levels tend to experience higher fentanyl overdose rates}.

First, the analysis of the **relationship between education levels and fentanyl overdose rates** reveals a weak negative correlation (correlation coefficient of around -0.2). States in the top quartile in terms of education (over 35% of adults with a bachelor's degree or higher) have an average overdose rate of 28.5 per 100,000 inhabitants, compared with 36.2 per 100,000 in states in the bottom quartile.

Moreover, examination of **regional disparities** shows that the Western region has the highest median and the greatest variability in total fentanyl seizures, suggesting its potential role as a major point of entry or transit for illicit drug trafficking.

After reviewing our research objectives, we will now assess the extent to which our initial hypotheses have been confirmed by our results.

The hypothesis that \uline{states with lower levels of education have higher rates of fentanyl overdose} and seizures is partially supported; however, the relationship is **weak and not statistically significant**.

Additionally, the \uline{relationship between fentanyl seizures and overdoses isn't significantly more pronounced in less-educated states}, which **contradicts our initial hypothesis**.

Our study has several important limitations. \uline{Firstly}, we did not take into account certain potential confounding factors, notably access to healthcare, which could significantly influence overdose rates independently of education levels.

\uline{Secondly}, the lack of complete data for the year 2022 limited the scope of our analysis, potentially reducing the reliability of our conclusions.

\uline{Finally}, overdose statistics include various drugs, not just fentanyl, which may bias our specific understanding of the fentanyl crisis. To address these limitations, we **recommend** incorporating data on access to care in future analyses, extending the study period to obtain a more robust dataset, and seeking data sources that differentiate overdoses by drug type.
